Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-06 source/paper.tex

\begin{equation}
  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
\end{equation}
where $\hat{\rho}$ denotes the value of $\rho$ on the element
boundary. Summing over the domain gives
\begin{equation}
  \int_\Omega \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_\Omega \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_\Gamma \jump{\phi\MM{u}}\tilde{\rho} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
\end{equation}
where $\Omega$ is the domain, $\Gamma$ denotes the element facets,
$\tilde{\rho}$ denotes the value of $\rho$ on the upwind side of the
facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
across the facet, where the $+$ and $-$ indicate each side of the
facet.

\subsubsection{Transport of $\mathbb{V}_\theta$ fields}
\label{subsec: theta transport}
The compressible Boussinesq and compressible Euler equation sets both
contain a prognostic equation that is the advective form of the
transport equation for a variable that is represented by a function
that is continuous in the vertical direction but discontinuous in the
horizontal direction. This means that the gradient of the field
required in the transport term is not well defined. As described in
\citet{yamazaki2017vertical}, a stable discretisation of this equation
can be constructed by using a combination of the above upwinding
strategy in the discontinuous horizontal direction and a
streamline-upwind Petrov-Galerkin (SUPG) method
\citep{brooks1982streamline} in the continuous vertical dirction. The
SUPG method replaces the finite element test function $\gamma$ with
$\gamma + \tau \MM{u}\cdot\nabla\gamma$ where $\tau$ is a
stabilisation parameter, although here we just require the vertical
part of the gradient term. This has the effect of suppressing spurious
oscillations associated with the continuous finite element
approximation of the transport term. The stabilisation parameter
$\tau$ is set to be $\Delta t/\sqrt{15}$ by default in Gusto,
following \citet{raymond1976selective}.

>>>>

\begin{equation}
  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
\end{equation}
where $\hat{\rho}$ denotes the value of $\rho$ on the element
boundary. Summing over the domain gives
\begin{equation}
  \int_\Omega \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_\Omega \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_\Gamma \jump{\phi\MM{u}}\tilde{\rho} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
\end{equation}
where $\Omega$ is the domain, $\Gamma$ denotes the element facets,
$\tilde{\rho}$ denotes the value of $\rho$ on the upwind side of the
facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
across the facet, where the $+$ and $-$ indicate each side of the
facet.

\subsubsection{Transport of $\mathbb{V}_\theta$ fields}
\label{subsec: theta transport}
The compressible Boussinesq and compressible Euler equation sets both
contain a prognostic equation that is the advective form of the
transport equation for a variable that is represented by a function
that is continuous in the vertical direction but discontinuous in the
horizontal direction. This means that the gradient of the field
required in the transport term is not globally defined. As described
in \citet{yamazaki2017vertical}, a stable discretisation of this
equation can be constructed by using a combination of the above
upwinding strategy in the discontinuous horizontal direction and a
streamline-upwind Petrov-Galerkin (SUPG) method
\citep{brooks1982streamline} in the continuous vertical direction. The
SUPG method replaces the finite element test function $\gamma$ with
$\gamma + \tau \MM{u}\cdot\nabla\gamma$ where $\tau$ is a
stabilisation parameter, although here we just require the vertical
part of the gradient term. This has the effect of suppressing spurious
oscillations associated with the continuous finite element
approximation of the transport term. The stabilisation parameter
$\tau$ is set to be $\Delta t/\sqrt{15}$ by default in Gusto,
following \citet{raymond1976selective}.
